---
title: "temporalCPSA Overview"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
    css: temporalCPSA-vignette.css
vignette: >
  %\VignetteIndexEntry{temporalCPSA Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## What Is CPSA?

Cross-Population Survival Analysis (CPSA) is a framework for assessing
prognostic molecular trajectories across complementary patient cohorts.

CPSA estimates temporal molecular patterns (TMPs) in deeply profiled discovery
cohorts, evaluates their clinical relevance in external clinically annotated
populations, and validates these prognostic patterns back in the discovery
cohort. By leveraging age-associated molecular structure rather than static
cohort-specific associations alone, CPSA supports cross-population
prioritization of prognostic signals that may reflect conserved biological
dynamics across heterogeneous datasets.

## What Can This Package Be Used For?

- Reproducing analyses from Tignor et al., *Proteogenomic analysis of pediatric
  and AYA high-grade glioma*, using the companion
  [Wang Lab manuscript repository](https://github.com/WangLab-MSSM/Pediatric-AYA-high-grade-glioma).
- Using deeply profiled CPTAC or related multi-omic cohorts to inform survival
  modeling in external clinically annotated populations with sparse or absent
  molecular data.
- Leveraging temporal molecular structure to identify prognostic biomarkers
  whose outcome associations reflect conserved biological dynamics across
  complementary patient cohorts.
- Modeling age-associated molecular trajectories to facilitate comparison of
  age-related molecular dynamics within and between cohorts, including tumor
  versus normal and male versus female tissue comparisons.
- Exploration of age structure within cohorts based on clustering of molecular
  abundance profiles across diverse multi-omic data types.

## General CPSA Workflow

The core workflow starts with a feature-by-sample matrix and a reference-cohort
clinical table containing survival time, event status, age class, and
adjustment covariates.

Default arguments in several trajectory and CPSA helpers reflect the manuscript
analysis settings; users should review and modify these settings for other
cohorts, molecular platforms, or modeling designs.

## Estimating Temporal Molecular Trajectories

The basic trajectory-estimation helper is
`ageTMP_predict_tumor_trajectory_matrix()`. It fits age-associated molecular
trajectories in a deeply profiled cohort and predicts those trajectory values
at the ages of a target cohort. The target cohort can be the same cohort, an
external clinically annotated cohort, or an age grid used for visualization.

```{r trajectory-estimation}
library(temporalCPSA)

tumor_matrix <- your_feature_by_sample_matrix
discovery_clinical <- your_discovery_clinical_table
target_clinical <- your_target_clinical_table

trajectory_fit <- ageTMP_predict_tumor_trajectory_matrix(
  tumor_mat = tumor_matrix,
  tumor_metadata = discovery_clinical,
  prediction_metadata = target_clinical,
  tumor_sample_col = "sample_id",
  tumor_age_col = "age",
  tumor_sex_col = "sex",
  prediction_sample_col = "sample_id",
  prediction_age_col = "age",
  prediction_sex_col = "sex",
  span = 0.75,
  prediction_scope = "all_samples",
  return_trajectory = TRUE
)

trajectory_matrix <- trajectory_fit$matrix
trajectory_long <- trajectory_fit$trajectory
```

For tumor-versus-normal/reference trajectory comparisons, use
`ageTMP_compare_normal_tumor_trajectory()` when both a tumor cohort and a
normal/reference cohort are available:

```{r normal-tumor-trajectory}
trajectory_comparison <- ageTMP_compare_normal_tumor_trajectory(
  tumor_mat = tumor_matrix,
  normal_mat = normal_reference_matrix,
  tumor_metadata = discovery_clinical,
  normal_metadata = normal_reference_clinical,
  tumor_sample_col = "sample_id",
  normal_sample_col = "sample_id",
  tumor_age_col = "age",
  normal_age_col = "age",
  tumor_sex_col = "sex",
  normal_sex_col = "sex",
  features = rownames(tumor_matrix)[1:25],
  span = 0.75
)
```

```{r cpsa-spec}
library(temporalCPSA)

feature_matrix <- your_feature_matrix
survival_data <- your_survival_data

spec <- ageTMP_cpsa_spec(
  time_col = "days",
  event_col = "os.status",
  age_class_col = "age_class",
  age_reference = "ADULT",
  strata = c("PED", "ADO", "YA"),
  base_covariates = c("age", "Grade", "Cortical", "Midline"),
  adjustment_covariates = c("TP53_mut", "ATRX_mut", "H33A_mut", "ATM_mut")
)

fit <- ageTMP_fit_reference_cpsa(
  feature_matrix = feature_matrix,
  survival_data = survival_data,
  spec = spec
)
```

For a standard Cox model without age-class interaction terms, set
`age_class_col`, `age_reference`, and `strata` to `NULL`:

```{r standard-cox}
standard_spec <- ageTMP_cpsa_spec(
  age_class_col = NULL,
  age_reference = NULL,
  strata = NULL,
  include_all_combined_test = FALSE,
  base_covariates = c("age", "Grade"),
  adjustment_covariates = character()
)

standard_fit <- ageTMP_fit_reference_cpsa(
  feature_matrix,
  survival_data,
  spec = standard_spec
)
```

## Trajectory Dynamic-Range Screening

When CPSA is applied to age-dependent trajectory features, users should screen
out features with minimal trajectory dynamic range before interpreting survival
associations.

```{r trajectory-sd}
trajectory_sd <- ageTMP_rank_trajectory_sd(
  tumor_trajectory,
  feature_col = "feature",
  value_col = "fit",
  group_cols = "trajectory_stratum"
)

fit <- ageTMP_fit_reference_cpsa(
  feature_matrix,
  survival_data,
  spec = spec,
  trajectory_sd = trajectory_sd,
  trajectory_sd_min = 0.15,
  trajectory_sd_group_cols = "trajectory_stratum",
  trajectory_sd_keep = "any_group"
)
```

`trajectory_sd_group_cols` identifies the column or columns defining trajectory
strata for dynamic-range screening. In the manuscript analyses this was often
sex, but the filter itself is generic. With `trajectory_sd_keep = "any_group"`,
a feature is retained if it is dynamic in at least one trajectory stratum.

The CPSA model engine computes raw p-values and, by default, Benjamini-Yekutieli
FDR columns with `stats::p.adjust(..., method = "BY")`. Because CPSA is often
applied across correlated molecular features and trajectory-derived scores,
users should inspect empirical test-statistic distributions and consider
whether additional calibration is appropriate for their modeling design.

In the manuscript reference-cohort analyses, protein and RNA significance calls
used a separate local-FDR procedure on transformed p-values. That procedure is
documented in the manuscript figure-generation scripts and is not part of the
general package API.

## Optional Age-Class Exploration

Age-class estimation is optional. Users may supply age classes, model age
continuously, or use `temporalCPSA` diagnostics to explore whether AD-TMP
structure supports age-contiguous intervals.

In the manuscript, age classes were chosen by considering both temporal
molecular clustering and conventional age distributions. In `temporalCPSA`, we
implement a data-driven diagnostic approach: candidate classes are constrained
to ordered age intervals, while the AD-TMP matrix determines which intervals are
best supported.

```{r age-class-diagnostic}
clinical <- ageTMP_load_discovery_clinical("data")
feature_matrices <- ageTMP_load_feature_matrix(
  "data",
  modality = c("protein", "rna", "phospho")
)

tmp_matrices <- lapply(feature_matrices, function(feature_matrix) {
  ageTMP_predict_tumor_trajectory_matrix(
    tumor_mat = feature_matrix,
    tumor_metadata = clinical,
    prediction_metadata = clinical,
    tumor_sample_col = "id",
    tumor_age_col = "cDisc_age",
    tumor_sex_col = "cDisc_Gender",
    prediction_sample_col = "id",
    prediction_age_col = "cDisc_age",
    prediction_sex_col = "cDisc_Gender",
    return_trajectory = FALSE
  )$matrix
})

combined_tmp <- ageTMP_combine_tmp_matrices(tmp_matrices)

fit <- ageTMP_segment_age_classes(
  tmp_matrix = combined_tmp,
  sample_age = setNames(clinical$cDisc_age, clinical$id),
  k = 5,
  min_n = 10,
  labels = c("PED", "ADO", "YA", "ADULT", "SEN")
)

fit$cutpoints
fit$segment_summary
```

To review the ordered segmentation landscape, draw the green diagnostic heatmap
and overlay optimizer-selected, manuscript-style, or investigator-selected
cutpoints.

```{r segment-diagnostic}
diagnostic <- ageTMP_segment_diagnostic_matrix(
  tmp_matrix = combined_tmp,
  sample_age = setNames(clinical$cDisc_age, clinical$id),
  k_values = 2:8,
  min_n = 10
)

ageTMP_plot_segment_diagnostic(
  diagnostic,
  selected_k = 5,
  class_labels = c("PED", "ADO", "YA", "ADULT", "SEN"),
  suggested_cutpoints = c(15, 26, 40, 62),
  suggested_labels = c("PED", "ADO", "YA", "ADULT", "SEN")
)
```

## Manuscript Reproduction

Paper figure-generation scripts live in the companion manuscript repository:
[`WangLab-MSSM/Pediatric-AYA-high-grade-glioma`](https://github.com/WangLab-MSSM/Pediatric-AYA-high-grade-glioma).

Manuscript-specific wrappers such as `ageTMP_build_sa_protein_cdisc()` are kept
for direct table reproduction because the published table columns use the
`cdisc` label. Preferred public-facing functions use reference-cohort language:

```{r manuscript-reproduction}
clinical <- ageTMP_load_reference_clinical("data")
mutation <- ageTMP_load_reference_mutation("data")
feature_matrix <- ageTMP_load_reference_protein_matrix("data")

survival_data <- ageTMP_prepare_reference_cpsa_survival(
  clinical = clinical,
  mutation = mutation,
  protein_sample_ids = colnames(feature_matrix),
  sex = "Female",
  mutation_na = "preserve"
)
```

## Reproducibility Principle

Reproducible workflows should read directly from documented source files
whenever possible and avoid hidden `.RData` objects or manually generated
intermediate TSV files. Serialized package data are reserved for documented
external reference data that are part of the analysis provenance.
